Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-07-12 analysis/paper/paper.qmd
editing methods section

In this study, we demonstrate a novel approach to investigating the technological status of spheroids with a case study from the Middle Pleistocene site of Qianshangying in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group spheroids and subspheroids. We observed notable differences between these artifacts and multifacial cores or polyhedrons <!-- BM: not sure what this next phrase means, I recommend you delete or edit to clarify:  "which are stone artifacts through typical flaking strategies" --> and hypothesized distinct reduction sequences. Our goal was to quantitatively evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes that were distinct from the multifacial cores in the assemblage. Our results contribute the debate about whether the concept and manufacture of spheroids represent a distinct technological tradition, separate from conventional flake-based strategies. To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying. Our approach includes the first open source and reproducible application of spherical harmonic analysis to a Paleolithic assemblage from East Asia. These analyses help clarify the technological strategy behind spheroid production at Qianshangying, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

Raw materials at Qianshangying. were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively). It seems that spheroids show no strong raw material preference. Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert. However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores. (pie chart in need?) <!-- BM: in your R console, please run ?pie then read the text under the 'Note' heading to see what I and many other people think about pie charts ;) -->

```{r, echo=FALSE}
#| label: fig-core-spheroid-size
#| fig-cap: "box plot showing size of cores and spheroids"
#| 
# 读取表格
library(tidyverse)
library(readxl)
library(here)
library(dplyr)

@@

# 画图
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(
    title = "Comparison of Core, Spheroid, and Multi-poly Core Dimensions",
    x = "Dimension",
    y = "Measurement (mm)",
    fill = "Artifact"
  ) +
  scale_fill_manual(values = c("Core" = "#66c2a5", 
                               "Spheroid" = "orange", 
                               "Multi-poly Core" = "lightgray")) +
  theme_minimal(base_size = 14)

@@

In terms of dimensions, the average length, width, thickness, and weight of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively. The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids.Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

In this study, the 3d models of the cores were collected and used for morphometric and technological analysis. Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio \[version number\]. The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/./xxxx/.

@@

Based on typological study, we identify some faceted sub-spheroids/spheroids in QSY site. To minimize subjective judgment inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to evaluate the degree of difference between the archaeological spheroids and a perfect sphere. In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.

These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact. We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies. Furthermore, we compared the results of the typological and our new morphometric analyses to investigate the validity of the traditional typological approach.

@@

To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research. The index is calculated using the following formula:

@@

Where V is the volume and A is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. While values closer to 1 generally indicate a more spherical geometry, sphericity is calculated solely from volume and surface area. As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to angular features or surface roughness. Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere. We used Python to derive surface area and volume measurements from 3D mesh models, and before that all models were decimated to a consistent face count and subjected to unit normalization to ensure comparability across artifacts.

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025]. Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

>>>>

In this study, we demonstrate a novel approach to investigating the technological status of spheroids with a case study from the Middle Pleistocene site of Qianshangying in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group spheroids and subspheroids. We observed notable differences between these artifacts and multifacial cores or polyhedrons <!-- BM: not sure what this next phrase means, I recommend you delete or edit to clarify:  "which are stone artifacts through typical flaking strategies" --> and hypothesized distinct reduction sequences. Our goal was to quantitatively evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes that were distinct from the multifacial cores in the assemblage. To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying. Our results contribute the debate about whether the concept and manufacture of spheroids represent a distinct technological tradition, separate from conventional flake-based strategies. Our approach includes the first open source and reproducible application of spherical harmonic analysis to an archaeological stone artefact assemblage. These analyses help clarify the technological strategy behind spheroid production at Qianshangying, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

Raw materials at Qianshangying. were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. Lava dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively). Spheroids show no strong raw material preference. Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert. However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores. (pie chart in need?) <!-- BM: in your R console, please run ?pie then read the text under the 'Note' heading to see what I and many other people think about pie charts ;) -->

```{r, echo=FALSE}
#| label: fig-core-spheroid-size
#| fig-cap: "Box plot comparing of core, spheroid, and multifacial core dimensions"
#| 
# 读取表格
library(tidyverse)
library(readxl)
library(here)
library(dplyr)

@@

# 画图
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Type"
  ) +
  scale_fill_manual(values = c("Core" = "#66c2a5", 
                               "Spheroid" = "orange", 
                               # BM: what is 'Multi-poly'? we don't have this word anywhere else in the text
                               "Multifacial Core" = "lightgray")) +
  theme_minimal(base_size = 14)

@@

In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively. The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

We generated 3d models of the cores for morphometric and technological analysis. Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio <!-- BM: add version number -->. The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/./xxxx/. <!-- BM: love to see this! -->

@@

In our typological study we identified some faceted sub-spheroids/spheroids in the Qianshangying assemblage. To minimize subjective judgments that are inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to our 3d models to evaluate the degree of difference between the archaeological spheroids and a perfect sphere. In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.

These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact. We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies. Furthermore, we compared the results of the typological study and our new morphometric analyses to investigate the validity of the traditional typological approach.

@@

To evaluate the degree to which the artifacts approximate a perfect sphere, we employed a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research. The index is calculated using the following formula:

@@

Where V is the volume and A is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. While values closer to 1 generally indicate a more spherical geometry, sphericity is calculated solely from volume and surface area. As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to angular features or surface roughness. Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere. We used Python to decimate to a consistent face count and conduct a unit normalization to ensure comparability across artifacts, and then derived surface area and volume measurements from these processed 3D mesh models.

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used the closed-source software Artefact3-D to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. @Noshita2025 . Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.
